Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-28 paper.tex
Up links
Update results, method and render AL

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{threeparttable}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

Perception of economic inequality has been associated with important
outcomes such as redistributive preferences, justification of
inequality, and legitimacy of the economic system (Castillo et al., 2022
{[}acá citar gente de Granada{]}). Research shows that greater perceived
inequality fosters stronger preferences for redistribution, regardless
of actual inequality levels
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}). However, other findings suggest that heightened
perceptions of income gaps between high- and low-paying occupations may
lead to greater justification of inequality
(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
this regard concerns market justice, which denotes the degree to which
individuals consider just that social goods and services (e.g., health,
pensions, education) are allocated according to individual contribution,
competition, and ability to pay
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
reflect the belief that the market promotes procedural
fairness---equality of opportunity---so that outcomes result from
individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
justice preferences are stronger in countries with higher private
spending on services than in those with more comprehensive welfare
systems. From this perspective, perceptions of economic inequality may
be key, as lower perceived gaps can reinforce acceptance of free-market
systems and justify existing inequalities.

Meritocracy posits that inequality is an inherent feature of societies
but can be legitimized through fairer principles, such as effort and
talent (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
inequality since these beliefs are associated with attitudes that
maintain the legitimization of social differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
like Chile, where the distribution of goods and services is
predominantly governed by market logics strongly introduced during the
military dictatorship (1973-1989)
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
play a crucial role in the acceptance of social inequalities.
Incorporating the perception of meritocracy in this literature allows
for an understanding of how attitudes toward inequality are shaped not
only by the perception of economic gaps, but also by beliefs about how
these gaps are justified. Furthermore, these beliefs are often
consolidated from an early age, reinforced by institutions promoting
values such as effort and individual skills to climb socially
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

@@

\subsection{Justification of market
inequality}\label{justification-of-market-inequality}

@@

Empirical studies have developed different strategies to capture and
operationalize market justice preferences. A common approach in the
literature is to gauge attitudes toward the legitimacy of inequality in
specific domains, especially when linked to income differences. Kluegel
and Smith (\citeproc{ref-kluegel_beliefs_1981}{1981}) and later Kluegel
et al. (\citeproc{ref-kluegel_legitimation_1999}{1999}) laid the
groundwork for measuring beliefs in market justice through survey items
that assess whether people support economic inequality when it reflects
personal effort and skill. Over time, this approach has been extended
beyond income to include other market-mediated outcomes, such as
education, healthcare, or pensions. For example, Knesebeck et al.~(2016)
and Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020})
examine whether citizens consider it fair that individuals with higher
incomes can access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Some research, like that of Lindh
(\citeproc{ref-lindh_public_2015}{2015}) and Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}), uses these indicators
in combination to measure broader preferences for market-based
distribution in welfare systems. More recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These different operationalizations all aim to
capture the extent to which individuals accept inequality when it is
framed as a reflection of market outcomes.

@@

@@

Among the different approaches to measuring perceived inequality, one of
the most widely used is the estimation of wage gaps between occupational
extremes, such as between a CEO and a manual worker. This type of item
provides a concrete frame that respondents can relate to more easily
than abstract questions about national income distribution {[}Castillo
et al. (\citeproc{ref-castillo_percepcion_2019}{2012}); Easterbrook
(\citeproc{ref-easterbrook_social_2021}{2021}); Willis et al., 2015{]}.
While it enables the estimation of inequality using simple heuristics,
this method is not without challenges. For instance, people often lack
reliable knowledge about the earnings of those at the top of the income
ladder, which leads to high variability in responses and the use of
biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
political
attitudes(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
et al., 2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz,
2019}), making them a valuable tool for understanding public responses
to economic disparities.

Another relevant critique of this approach lies in its conflation of
different psychological constructs. Many surveys assess perceived
inequality through Likert-type items that ask respondents to agree or
disagree with statements, such as ``income differences are too large,''
which captures general concern or discomfort rather than a specific
perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).These items mix cognitive estimations with affective evaluations,
complicating the interpretation of what respondents perceive versus what
they morally reject. As a result, the conceptual clarity between
perceived inequality and inequality aversion remains blurred in many
empirical studies. To address this limitation, recent work has
emphasized the need to distinguish between absolute and comparative
measures, as well as between ideal and actual estimates of economic gaps
{[}García-Sánchez \& De Carvalho
(\citeproc{ref-garcia-sanchez_creencias_2022}{2022}); Auspurg et al.,
2017{]}.

Importantly, perceptions of inequality have been linked to attitudes
about market justice---the belief that goods and services should be
allocated according to individual contribution, competition, or
willingness to pay (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
indicates that lower perceived inequality can reinforce support for
market-based distributive arrangements by suggesting that the system is
fair and that outcomes reflect effort and ability
(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
is perceived as excessive or structurally determined, individuals could
question the legitimacy of market justice and become more supportive of
redistributive policies
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

@@

Meritocracy constitutes a central ideological framework for legitimizing
social inequality. Rooted in the belief that rewards and positions
should be allocated based on individual effort and talent, meritocracy
operates as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
system in which merit-based stratification becomes a new form of
inequality. However, over time, meritocracy has been widely adopted in
many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
perspective, the belief in meritocracy is more than a cognitive
assessment; it reflects a moral lens through which individuals interpret
inequality. People who believe that success results from hard work and
talent are more likely to view social and economic disparities as
legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
Conversely, if they see outcomes as driven by luck, social origin, or
systemic barriers, inequality is more likely to be perceived as unjust.
This distinction becomes crucial in societies with persistent structural
inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they personally
endorse meritocratic principles. This distinction aligns with recent
findings indicating that individuals distinguish between how merit is
perceived in society and how it should ideally operate, which in turn
shapes their preferences for redistribution and justice
(Tejero-Peregrina et al., 2025). Meritocratic beliefs serve as symbolic
justifications for unequal outcomes, particularly when access is
stratified by income or social opportunity. Prior studies have shown
that individuals who perceive higher levels of meritocracy tend to
express stronger support for unequal distributions that reflect market
outcomes (\citeproc{ref-castillo_percepcion_2019}{Castillo et al.,
2012}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}).

In addition to influencing individual attitudes toward inequality,
meritocratic beliefs can contribute to social division and the
stigmatization of disadvantaged groups. Recent research has demonstrated
that exposure to meritocratic narratives can reinforce the belief that
poverty results from individual failings rather than systemic
conditions, reducing support for redistributive measures and increasing
the stigmatization of the poor (Hoyt \& Burnette, 2023). This reinforces
negative stereotypes and reduces empathy toward individuals from lower
socioeconomic backgrounds. Moreover, Busemeyer, Abrassart, and Nezi
(2019) argue that meritocratic narratives can serve as feedback
mechanisms that shape public opinion and well-being by framing
individuals' understanding of welfare outcomes as deserved or undeserved
within existing institutional structures. This psychological mechanism
highlights the normative power of meritocracy in stabilizing unequal
systems by shaping political attitudes and personal perceptions of
success and failure.

\subsection{This study (hypotheses)}\label{this-study-hypotheses}

@@

\emph{Dependent variable}

@@

\emph{Independent variables}

\textbf{Perception of economic inequality}: The main independent
variable refers to the perception of economic inequality, measured
through the perceived wage gap, which serves as an absolute indicator of
inequality perceptions (\citeproc{ref-castillo_perception_2022}{Castillo
et al., 2022}). This measure is derived from the gap between the
perceived salaries of jobs at opposite ends of the occupational
hierarchy. Specifically, it relies on the logarithm of the division
between the perceived salary of a large-company president and that of an
unskilled worker (\citeproc{ref-Castillo2011}{Castillo, 2011}). Higher
values of this term indicate a greater perception of economic inequality
between occupations located at the extremes of the status continuum.

\textbf{Justification of economic inequality}: This construct is
measured as the just earnings gap, an approach to measuring the
justification of economic inequality based on earning differences
considered just between high and low status occupations
(\citeproc{ref-Castillo2011}{Castillo, 2011}). Similarly to economic
inequality perception, this indicator relies on the logarithmic ratio
between the just salary of a large-company president and that of an
unskilled worker. Higher values of this term indicate a greater
justification of economic inequality.

\textbf{Meritocracy}: Meritocratic perception is operationalized through
two components: one addressing effort and another focusing on talent
(\citeproc{ref-young_rise_1962}{Young, 1962}). The item used to gauge
effort is: ``In Chile, people are rewarded for their efforts,'' while
the item for talent is: ``In Chile, people are rewarded for their
intelligence and skills.'' In both cases, respondents indicate their
level of agreement on a five-point Likert scale, ranging from 1
(``strongly disagree'') to 5 (``strongly agree'').

@@

Sociodemographic and attitudinal variables are included to control for
potential composition effects in the population. In terms of
sociodemographic characteristics, we incorporate per capita household
income quintile, educational level (1=Less than Universitary,
2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
been shown to influence market justice preferences significantly
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding attitudinal
variables, we include political identification (1=Left, 2=Center,
3=Right, 4=No identification) and subjective social status (ranging from
1 to 10) because they may confound the relationship between market
justice preferences and perceptions of inequality and meritocracy
(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).

@@

Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
preferences for healthcare, pensions, and education from 2016 to 2023.
Each year presents stacked percentage frequencies, and the flows between
them reflect opinion changes among the same individuals from one year to
the next, given that we are using panel data. For instance, of the
40.9\% who strongly disagreed with justifying inequality in healthcare
in 2019, around 24.3\% maintained that position in 2022, while the
remaining 16.6\% shifted toward other response categories---primarily
moving into disagreement rather than strong disagreement. Overall, more
than half of the respondents exhibit a high level of disagreement
(disagree + strongly disagree) with inequality in these three social
service areas over time. Despite this general pattern, recent waves show
a slight decrease in disagreement and a corresponding rise in support
for inequality. Specifically, in healthcare and education, although
disagreement remains substantial, agreement increased from 8.9\% and
7.6\% in 2019 to 12\% and 12.5\% in 2023, respectively. This shift is
most evident in pensions, where the combined agree/strongly agree
category grew by about 10 percentage points, from 16.9\% in 2016 to
27.8\% in 2023.

@@

\caption{\label{fig-alluvial}Change in the justification of inequality
in health, pensions and education over time (2016-2023)}

@@

Regarding the main variables of this study, Figure~\ref{fig-meanchange}
depicts their average changes over the years. We observe an increase in
the average level of market justice preferences in the most recent
waves, which begins in 2019. The highest average consistently appears
for perceived economic inequality, although this variable shows a
downward trend of roughly one point over time. Interestingly, while
perceptions of economic inequality declined in the latest measurements
(2022--2023), market justice preferences increased. With respect to
economic inequality justification, there is a steady decrease in average
levels over the years, registering the lowest mean compared to the other
variables. The meritocracy measures remain stable, though the perception
that individuals are rewarded for intelligence is slightly higher than
the perception that they are rewarded for their effort.

@@

\caption{\label{fig-meanchange}Change in the mean of market justice
preferences, economic inequality perception and justification, and
meritocracy (2016-2023)}

@@

\scalebox{1}{
\begin{threeparttable}
\begin{tabular}{l c c c c c c c c}

>>>>

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

Perception of economic inequality has been associated with important
outcomes such as redistributive preferences, justification of
inequality, and legitimacy of the economic system
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
shows that greater perceived inequality fosters stronger preferences for
redistribution, regardless of actual inequality levels
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}). However, other findings suggest that heightened
perceptions of income gaps between high- and low-paying occupations may
lead to greater justification of inequality
(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
this regard concerns market justice, which denotes the degree to which
individuals consider just that social goods and services (e.g., health,
pensions, education) are allocated according to individual contribution,
competition, and ability to pay
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
reflect the belief that the market promotes procedural
fairness---equality of opportunity---so that outcomes result from
individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
justice preferences are stronger in countries with higher private
spending on services than in those with more comprehensive welfare
systems. From this perspective, perceptions of economic inequality may
be key, as lower perceived gaps can reinforce acceptance of free-market
systems and justify existing inequalities.

Meritocracy posits that inequality can be legitimized through
distributive criteria such as effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
inequality as these beliefs are associated with attitudes that
legitimize status differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
like Chile, where the distribution of goods and services is
predominantly governed by market logics strongly introduced during the
military dictatorship (1973-1989)
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
play a crucial role in the acceptance of social inequalities.
Incorporating the perception of meritocracy in this literature allows
for an understanding of how attitudes toward inequality are shaped not
only by the perception of economic gaps, but also by beliefs about how
these gaps are justified. Furthermore, these beliefs are often
consolidated from an early age, reinforced by institutions promoting
values such as effort and individual skills to climb socially
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

@@

\subsection{The justification of market
inequality}\label{the-justification-of-market-inequality}

@@

Empirical studies have developed different strategies to capture and
operationalize market justice preferences. A common approach in the
literature is to gauge attitudes toward the legitimacy of inequality in
specific domains, especially when linked to income differences. Kluegel
and Smith (\citeproc{ref-kluegel_beliefs_1981}{1981}) and later Kluegel
et al. (\citeproc{ref-kluegel_legitimation_1999}{1999}) laid the
groundwork for measuring beliefs in market justice through survey items
that assess whether people support economic inequality when it reflects
personal effort and skills. Over time, this approach has been extended
beyond income to include other market-mediated outcomes, such as
education, healthcare, and/or pensions. For example, Von Dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Some research, like that of Lindh
(\citeproc{ref-lindh_public_2015}{2015}) and Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}), uses these indicators
in combination to measure broader preferences for market-based
distribution in welfare systems. More recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These different operationalizations all aim to
capture the extent to which individuals accept inequality when it is
framed as a reflection of market outcomes.

@@

Perceptions of inequality have been associated to attitudes about market
justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al., 1995};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research indicates that
lower perceived inequality can reinforce support for market-based
distributive arrangements by suggesting that the system is fair and that
outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
2011}). In contrast, when inequality is perceived as excessive or
structurally determined, individuals could question the legitimacy of
market justice and become more supportive of redistributive policies
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

@@

Among the different approaches to measuring perceived inequality, one of
the most widely used is the estimation of wage gaps between occupational
extremes, such as between a CEO and a manual worker. This type of item
provides a concrete frame that respondents can relate to more easily
than abstract questions about national income distribution
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). While it
enables the estimation of inequality using simple heuristics, this
method is not without challenges. For instance, people often lack
reliable knowledge about the earnings of those at the top of the income
ladder, which leads to high variability in responses and the use of
biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
political attitudes
(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz, 2019}),
making them a valuable tool for understanding public responses to
economic disparities.

Another relevant critique of this approach lies in its conflation of
different psychological constructs. Many surveys assess perceived
inequality through Likert-type items that ask respondents to agree or
disagree with statements, such as ``income differences are too large,''
which captures general concern or discomfort rather than a specific
perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). These items mix cognitive estimations with affective
evaluations, complicating the interpretation of what respondents
perceive versus what they morally reject. As a result, the conceptual
clarity between perceived inequality and inequality aversion remains
blurred in many empirical studies. To address this limitation, recent
work has emphasized the need to distinguish between absolute and
comparative measures, as well as between ideal and actual estimates of
economic gaps (\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
Carvalho, 2022}).

@@

Meritocracy constitutes a central ideological framework for legitimizing
different types od social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
system in which merit-based stratification becomes a new form of
inequality. However, over time, meritocracy has been widely supported in
many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
perspective, the belief in meritocracy is more than a cognitive
assessment; it reflects a moral lens through which individuals interpret
inequality. People who believe that success results from hard work and
talent are more likely to view social and economic disparities as
legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
Conversely, if they see outcomes as driven by luck, social origin, or
systemic barriers, inequality is more likely to be perceived as unjust.
This distinction becomes crucial in societies with persistent structural
inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

@@

These dimensions of meritocracy reflect how respondents perceive
society's distributive logic, regardless of whether they personally
endorse meritocratic principles. This distinction aligns with recent
findings indicating that individuals distinguish between how merit is
perceived in society and how it should ideally operate, which in turn
shapes their preferences for redistribution and justice
(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}) Meritocratic beliefs serve as symbolic justifications for unequal
outcomes, particularly when access is stratified by income or social
opportunity. Prior studies have shown that individuals who perceive
higher levels of meritocracy tend to express stronger support for
unequal distributions that reflect market outcomes
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

In addition to influencing individual attitudes toward inequality,
meritocratic beliefs can contribute to social division and the
stigmatization of disadvantaged groups. Recent research has demonstrated
that exposure to meritocratic narratives can reinforce the belief that
poverty results from individual failings rather than systemic
conditions, reducing support for redistributive measures and increasing
the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
et al., 2023}). This reinforces negative stereotypes and reduces empathy
toward individuals from lower socioeconomic backgrounds. Moreover,
Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argue
that meritocratic narratives can serve as feedback mechanisms that shape
public opinion and well-being by framing individuals' understanding of
welfare outcomes as deserved or undeserved within existing institutional
structures. This psychological mechanism highlights the normative power
of meritocracy in stabilizing unequal systems by shaping political
attitudes and personal perceptions of success and failure.

\subsection{This study}\label{this-study}

Building upon the previous literature, this study proposes that
attitudes toward market justice are shaped by a dynamic interplay
between individuals' perceptions of economic inequality and their
beliefs in meritocracy. Specifically, we argue that both perceptions
independently and interactively influence the extent to which
individuals endorse market-based distributions of social goods and
services in Chile.

First, consistent with previous findings, we expect that a lower
perception of economic inequality will be associated with stronger
market justice preferences. When individuals perceive smaller economic
gaps, they are more likely to view market mechanisms as fair and
legitimate, reinforcing acceptance of outcomes based on competition and
ability to pay. Conversely, a heightened perception of inequality may
erode confidence in market fairness, weakening support for market-based
distribution. This relationship is particularly relevant in the context
of Chile, where the neoliberal economic model has been a dominant force
shaping public attitudes toward inequality and justice.

Second, stronger meritocratic beliefs are expected to be positively
associated with market justice preferences. Individuals who believe that
effort and talent primarily determine success are more likely to justify
unequal outcomes and endorse the notion that markets allocate resources
fairly according to individual merit. This aligns with the idea that
meritocratic beliefs serve as a moral framework that legitimizes
market-based inequalities, as individuals perceive the system as just
when they believe that rewards are based on individual merit. This is
particularly relevant in contexts where neoliberal ideologies dominate,
as they often emphasize individual responsibility and competition as the
basis for social order.

Third, we propose that perceptions of meritocracy and perceptions of
economic inequality interact in shaping market justice attitudes.
Specifically, we argue that the legitimizing effect of meritocratic
beliefs on market justice preferences is moderated by perceptions of
economic inequality: when perceived inequality is low, the positive
association between meritocratic beliefs and market justice preferences
will be stronger. However, when perceived inequality is high, this
association will weaken, as greater awareness of large economic gaps may
challenge the view that outcomes are purely merit-based.

Additionally, this study examines whether these relationships vary over
time in response to significant social and political events. In
particular, we explore whether the political outburst of 2019 and the
subsequent constitutional processes in Chile have altered how
perceptions of inequality and meritocracy shape market justice
attitudes. We expect that after these critical events, perceptions of
inequality may have a stronger negative effect on market justice
preferences, reflecting increased societal questioning of market-based
allocation mechanisms.

Based on these arguments, we propose the following hypotheses:

\(H1\): Greater perceived economic inequality is associated with
stronger market justice preferences.

\(H2\): Stronger meritocratic beliefs are associated with stronger
market justice preferences.

\(H3\): The positive association between meritocratic beliefs and market
justice preferences is moderated by perceived economic inequality;
specifically, this association is weaker when perceived economic
inequality is high.

\(H4\): The effects described in H1--H3 are attenuated after major
social mobilizations (2019--2022), reflecting increased critical views
of the market's role in allocating social goods.

@@

@@

\textbf{Perception of economic inequality}: The main independent
variable refers to the perception of economic inequality, measured
through the perceived wage gap.
(\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
\citeproc{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
\citeproc{ref-hadler_why_2005}{Hadler, 2005}). This measure is derived
from the salary gap between the perceived salaries of jobs at opposite
ends of the occupational hierarchy. Specifically, it relies on the
division between the perceived salary of a large-company president and
that of an unskilled worker (\citeproc{ref-Castillo2011}{Castillo,
2011}). Higher values of this term indicate a greater perception of
economic inequality between occupations located at the extremes of the
status continuum. This measure includes a logarithmic term in order to
adjust income magnitudes (usually fewer cases with high income):

\[
\text{perceived wage gap} = \log_{10}\left(\frac{\text{perceived salary of a large-company president}}{\text{perceived salary of an unskilled worker}}\right)
\]

\textbf{Perception of Meritocracy}: this variable is operationalized
through two components, namely effort and talent
(\citeproc{ref-young_rise_1962}{Young, 1962}). The item used to gauge
effort is: ``In Chile, people are rewarded for their efforts,'' while
the item for talent is: ``In Chile, people are rewarded for their
intelligence and skills''. In both cases, respondents indicate their
level of agreement on a five-point Likert scale, ranging from 1
(``strongly disagree'') to 5 (``strongly agree'').

@@

Sociodemographic and attitudinal variables are included to control for
potential composition effects in the population. In terms of
sociodemographic characteristics, we incorporate per capita household
income quintile, educational level (1=Less than Universitary,
2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
been previously shown to influence market justice preferences
significantly (\citeproc{ref-castillo_socialization_2024}{Castillo et
al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
attitudinal variables, we include political identification (1=Left,
2=Center, 3=Right, 4=No identification) and subjective social status
(ranging from 1 to 10) as they may confound the relationship between
market justice preferences and perceptions of inequality and meritocracy
(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).

@@

Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
preferences for healthcare, pensions, and education from 2016 to 2023.
Each year presents stacked percentage frequencies, and the flows between
them reflect opinion changes among the same individuals from one year to
the next, given that we are using panel data. For instance, of the
40.8\% who strongly disagreed with justifying inequality in healthcare
in 2019, around 24.3\% maintained that position in 2022, while the
remaining 16.5\% shifted toward other response categories---primarily
moving into disagreement rather than strong disagreement. Overall, more
than half of the respondents exhibit a high level of disagreement
(disagree + strongly disagree) with inequality in these three social
service areas over time. Despite this general pattern, recent waves show
a slight decrease in disagreement and a corresponding rise in support
for inequality. Specifically, in healthcare and education, although
disagreement remains substantial, agreement (agree + strongly agree)
increased from 7.4\% and 7.2\% in 2019 to 13.1\% and 14.2\% in 2023,
respectively. This shift is most evident in pensions, where the combined
agree/strongly agree category grew by about 10 percentage points, from
16.9\% in 2016 to 28\% in 2023.

@@

\caption{\label{fig-alluvial}Change in the justification of inequality
in healthcare, pensions and education over time (2016-2023)}

@@

Regarding the main variables of this study, Figure~\ref{fig-meanchange}
depicts their average changes over the years. We observe an increase in
the average level of market justice preferences in the most recent
waves, which begins in 2019. The highest average consistently appears
for perceived economic inequality, although this variable shows a
downward trend of roughly one point over time. Interestingly, while
perceptions of economic inequality declined in the latest measurements
(2022--2023), market justice preferences increased. The meritocracy
measures remain stable, though the perception that individuals are
rewarded for intelligence is slightly higher than the perception that
they are rewarded for their effort.

@@

\caption{\label{fig-meanchange}Change in the mean of market justice
preferences, economic inequality perception, and meritocracy
(2016-2023)}

@@

\scalebox{0.79}{
\begin{tabular}{l c c c c c c c}
\hline
